---
title: Gaussian Process Regression
output:
   prettydoc::html_pretty:
    theme: leonids
    highlight: github

vignette: >
  %\VignetteIndexEntry{Gaussian Process Regression}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

<style>
body {
text-align: justify}
</style>

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(dpi=300, collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
knitr::opts_chunk$set(fig.align="center")
```

```{r message=FALSE, warning=FALSE}
library(MagmaClustR)
library(dplyr)
library(ggplot2)
```

## Classical pipeline

The overall pipeline for standard GP regression in *MagmaClustR* can be decomposed in 3 main steps: **training, prediction** and **plotting of results**.
The corresponding functions are:

-   [`train_gp()`](https://arthurleroy.github.io/MagmaClustR/reference/train_magma.html)
-   [`pred_gp()`](https://arthurleroy.github.io/MagmaClustR/reference/pred_gp.html)
-   [`plot_gp()`](https://arthurleroy.github.io/MagmaClustR/reference/plot_gp.html)

## Data

### Dataset format

Before using `train_gp()`, our dataset should present a particular format. It must contains those 2 columns:

*   `Input`: `numeric`,
*   `Output`: `numeric`. 

If we fail to ensure that the dataset satisfies those conditions, the algorithm will return an error. 

The data frame can also provide as many covariates (*i.e.* additional columns) as desired, with no constraints on the column names (except the name 'Reference' that should always be avoided, as it is used inside internal functions of the algorithm). These covariates are treated as additional input dimensions. 

### Example with swimming data

To explore the features of standard GP in *MagmaClustR*, we use the `swimmers` dataset provided by the French Swimming Federation (available [here](https://github.com/ArthurLeroy/MAGMAclust/blob/master/Real_Data_Study/Data/db_100m_freestyle.csv), and studied more thoroughly [here](https://link.springer.com/article/10.1007/s10994-022-06172-1) and [there](https://arxiv.org/abs/2011.07866)).

Our goal is to model the progression curves of swimmers in order to forecast their future performances. Thus, we randomly select a female swimmer from the dataset; let's give her the fictive name *Michaela* for the sake of illustration.

The `swimmers` dataset contains 4 columns: `ID`, `Age`, `Performance` and `Gender`.
Therefore, we first need to change the name and type of the columns, and remove `Gender` before using `train_gp()`.

```{r}
Michaela <- swimmers %>% filter(ID == 1718) %>% 
  select(-Gender) %>% 
  rename(Input = Age, Output = Performance)
```

We display Michaela's performances according to her age to visualise her progression from raw data.

```{r fig.asp = 0.65, fig.width = 7, out.width = "80%"}
ggplot() +
  geom_point(data = Michaela,
       mapping = aes(x=Input,y=Output),
       size = 3,
       colour = "black") +
  theme_classic()
```

## Fit a GP on Michaela's data points

### Training

To obtain a GP that best fits our data, we must specify some parameters:

*   `prior_mean`: if we assume no *prior* knowledge about the 100m freestyle, we can decide to leave the default value for this parameter (*i.e.* zero). However, if we want to take expert advice into account, we can modify the value of `prior_mean` accordingly.

*   `kern`: the relationship between observed data and prediction targets can be control through the covariance **kernel**.
Therefore, in order to correctly fit our data, we need to choose a suitable covariance kernel.
In the case of swimmers, we want a smooth progression curve for Michaela; therefore, we specify `kern = "SE"`.

The most commonly used kernels and their properties are covered in [the kernel cookbook](https://www.cs.toronto.edu/~duvenaud/cookbook/).
Details of available kernels and how to combine them in *MagmaClustR* are available in [`help train_gp`](https://arthurleroy.github.io/MagmaClustR/reference/train_gp.html). 


```{r}
set.seed(2)
model_gp <- train_gp(data = Michaela,
                     kern = "SE",
                     prior_mean = 0)
```

Thanks to `train_gp()`, we learn the model hyper-parameters from data, which can then be used to make prediction for Michaela's performances.

### GP prediction

The arguments `kern` and `mean` remain the same as above. We now need to specify:

*   the hyper-parameters obtained with `train_gp()` in `hp`;
*   the input values on which we want to evaluate our GP in `grid_inputs`. Here, we want to predict Michaela's performances until 15/16 years old, so we set `grid_inputs = seq(11, 15.5, 0.1)`.

See [`help pred_gp`](https://arthurleroy.github.io/MagmaClustR/reference/pred_gp.html) to get information about the other optional arguments.

```{r}
pred_gp <- pred_gp(data = Michaela,
                   kern = "SE",
                   hp = model_gp,
                   grid_inputs = seq(11,15.5,0.1),
                   plot = FALSE) 
```


Let us note that we could have called the `pred_gp()` function directly on data.
If we didn't use the `train_gp()` function beforehand, it is possible to omit the `hp` parameter; `pred_gp()`  would automatically learn hyperparameters with default settings and us them to make predictions afterwards.


### Plotting the results

Finally, we display Michaela's prediction curve and its 95% associated credibility interval thanks to the `plot_gp()` function. If we want to get a prettier graphic, we could specify `heatmap = TRUE` to create a heatmap of probabilities instead (note that it could be longer to display when evaluating on fine grids).

```{r fig.asp = 0.6, fig.width = 7, out.width = "80%"}
plot_gp(pred_gp = pred_gp,
        data = Michaela)
```


Close to Michaela’s observed data ($t \in [ 10, 14 ]$), the standard GP prediction behaves as expected: it accurately fits data points and the confidence interval is narrow.
However, as soon as we move away from observed points, the GP drifts to the **prior mean** and uncertainty increases significantly.



